---
title: "Error Codes and the Researcher's Data Dictionary"
vignette: >
  %\VignetteIndexEntry{Error Codes and the Researcher's Data Dictionary}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ntrs)
```

Neuropsychological data from NACC often contains special numeric codes that
don't represent actual test performance — a value of `88` for MOCATOTS doesn't
mean a score of 88, it means the test was not administered. Handling these
codes correctly is critical to avoid nonsensical results.

`ntrs` builds error-code awareness directly into its `npsych_scores` class so
that codes are validated on input, tracked through operations, and
automatically handled during standardization.

## The `rdd` object

`ntrs` ships with `rdd`, a list-based version of the NACC Researcher's Data
Dictionary. Each entry contains the valid range, special codes, and a short
description for a NACC variable:

```{r}
#| label: rdd-structure
names(rdd$MOCATOTS)

rdd$MOCATOTS$short_descriptor
rdd$MOCATOTS$range
rdd$MOCATOTS$codes
```

This is useful for quick lookups without needing to consult the PDF:

```{r}
#| label: rdd-examples
# What codes does TRAILA have?
rdd$TRAILA$codes

# What is the valid range for ANIMALS?
rdd$ANIMALS$range
```

The `rdd` object covers all variables in the demo dataset, not just those with
`npsych_scores` constructors:

```{r}
#| label: rdd-non-score
rdd$SEX$codes
rdd$RACE$codes
```

## Error codes vs. in-range codes

`npsych_scores` objects distinguish between two kinds of special codes:

- **Error codes** fall *outside* the valid score range. These represent
  situations where no valid score was obtained (e.g., test not administered,
  physical problem). They are replaced with `NA` by `remove_error_codes()`.

- **In-range codes** fall *within* the valid score range. These are unusual
  but can occur when a code value happens to be a plausible score.

For most tests, all codes are error codes (outside range):

```{r}
#| label: error-vs-inrange
moca <- MOCATOTS(c(25, 88, -4, 28))
moca

# 88 and -4 are both outside 0-30, so both are error codes
moca@codes
moca@range
```

`remove_error_codes()` identifies codes outside the range and replaces them
with `NA`:

```{r}
#| label: remove-error-codes
remove_error_codes(moca)
```

An exampel of a test with an in-range code is the Global CDR Score.^[In the official documentation for CDRGLOB, `99` is not listed, but this value shows up in the NACC data. It is therefore included as an error code in `CDRGLOB()`.]

```{r}
#| label: cdr-glob-example

cdr <- CDRGLOB(c(0, 0.5, 1, 2, 3, 99))
cdr

# all values are within the valid range, and all are codes
cdr@range
cdr@codes

# only 99 is removed
remove_error_codes(cdr)
```

Only codes that fall outside the range are removed — if a test happened to
have a code whose numeric value falls within the valid range, it would be
preserved by `remove_error_codes()`.

## Viewing code labels with `replace_codes()`

While `remove_error_codes()` is designed for analysis (replacing codes with
`NA`), `replace_codes()` is useful for inspection — it substitutes each code
value with its human-readable label:

```{r}
#| label: replace-codes
replace_codes(moca)
```

This is handy when exploring data to understand *why* values are missing.

## Validation on construction

When you create an `npsych_scores` object, the constructor validates that
every value is either within the declared range, matches a registered code, or
is `NA`. Values that don't fit any category are rejected:

```{r}
#| label: validation-error
#| error: true
# 50 is outside MOCATOTS range (0-30) and is not a registered code
MOCATOTS(c(25, 50))
```

This catches data-entry errors and miscoded values early, before they can
silently corrupt downstream analyses.

## Error codes during standardization

Both `std_using_norms()` and `std_using_regression()` call
`remove_error_codes()` internally before computing z-scores. This means error
codes automatically become `NA` in the standardized output — you don't need to
clean them manually:

```{r}
#| label: std-with-codes
scores <- MOCATOTS(c(25, 88, -4, 28, 30))

z <- std(scores, age = 72, sex = 1, educ = 16, race = 1)
z
```

The positions corresponding to error codes (`88` and `-4`) are `NA` in the
result, while valid scores are standardized normally.

## Inspecting codes for any test

You can check which codes a score type recognizes by examining the `codes`
property:

```{r}
#| label: inspect-codes
TRAILA()@codes

ANIMALS()@codes

REYTCOR()@codes
```

Many tests share a common set of codes (`-4` for "data not collected", `95`
through `98` for various problems). The NACC battery also uses `88` in some
tests for "not administered" or "not assessed, optional".

## Practical patterns

### Counting missing data by reason

Since `replace_codes()` turns numeric codes into descriptive labels, you can
use it to tabulate *why* data is missing:

```{r}
#| label: tabulate-codes
scores <- MOCATOTS(demo_data$MOCATOTS)
labeled <- replace_codes(scores)

# Count occurrences of each code label
table(labeled[!labeled %in% as.character(0:30)])
```

### Filtering to valid observations

A common pattern is to filter a dataset to rows where a particular test was
actually administered:

```{r}
#| label: filter-valid
df <- demo_data
df$MOCATOTS <- MOCATOTS(df$MOCATOTS)

# Keep only rows with valid (non-error-code, non-NA) MOCATOTS
valid_scores <- remove_error_codes(df$MOCATOTS)
df_valid <- df[!is.na(valid_scores), ]
nrow(df_valid)
```

### Checking multiple tests at once

When working with several tests, you might want to identify rows where *all*
tests of interest have valid data:

```{r}
#| label: complete-cases
df <- demo_data
df$MOCATOTS <- MOCATOTS(df$MOCATOTS)
df$ANIMALS <- ANIMALS(df$ANIMALS)
df$TRAILA <- TRAILA(df$TRAILA)

valid_moca <- !is.na(remove_error_codes(df$MOCATOTS))
valid_animals <- !is.na(remove_error_codes(df$ANIMALS))
valid_traila <- !is.na(remove_error_codes(df$TRAILA))

all_valid <- valid_moca & valid_animals & valid_traila
sum(all_valid)
```
