---
title: "New Standardization Methods"
vignette: >
  %\VignetteIndexEntry{New Standardization Methods}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ntrs)
```

`ntrs` ships with two standardization methods — `norms` and `regression` — but
the system is designed to be extended. Any S7 generic named `std_using_*` is
automatically discovered by `list_std_methods()` and can be called via `std()`.

This vignette walks through creating a new method from scratch, then discusses
how to package it as a standalone extension (using
[ntrsTscores](https://github.com/rmtrane/ntrsTscores) as a real-world example).


## Part 1: A toy method — `std_using_percentile`

Let's create a method that converts raw scores to percentile ranks using the
empirical CDF from a reference sample.

### Defining the generic

A standardization generic must:

1. Be named `std_using_<method>` (so `ntrs` can discover it)
2. Be an S7 generic dispatching on `scores`

```{r}
#| label: define-generic
std_using_percentile <- S7::new_generic(
  "std_using_percentile",
  dispatch_args = "scores"
)
```

### Implementing a method on `npsych_scores`

Implementing a method on the parent class `npsych_scores` means it applies to
**all** score types. This is the simplest approach when the logic is the same
across tests.

Our method will take a `ref_scores` argument — a numeric vector of reference
sample scores — and compute percentile ranks:

```{r}
#| label: method-npsych-scores
S7::method(std_using_percentile, npsych_scores) <- function(
  scores,
  ...,
  ref_scores
) {
  raw <- as.numeric(remove_error_codes(scores))
  ecdf_fn <- stats::ecdf(ref_scores)
  ecdf_fn(raw) * 100
}
```

That's it. The method is immediately available:

```{r}
#| label: list-methods
list_std_methods(MOCATOTS())
```

And callable via `std()`:

```{r}
#| label: std-percentile
moca <- MOCATOTS(c(22, 25, 28))
ref <- c(20, 22, 23, 24, 25, 25, 26, 27, 28, 29, 30)

std(moca, method = "percentile", ref_scores = ref)
```

Notice that our method doesn't use a `version` parameter. That's fine — `std()`
passes `version` only when it's non-NULL, and `list_std_methods()` doesn't
require version registration for methods without a `version` formal. Also, notice how `std_using_percentile()` removes error codes before computing percentiles. It is on the method implementer to decide how to handle error codes — the `remove_error_codes()` helper is available, and recommended, but needs to be explicitly called.


### Using the version registry (optional)

If you want your method to support multiple named versions (e.g., different
reference samples), add a `version` parameter and use the version registry.
This requires a few more steps:

1. Create an S7 class for your version data (inheriting from `std_version`)
2. Write a registration function
3. Use `get_version_data()` in your method

Let's extend our percentile method to support named reference datasets.

**Step 1: Create an S7 class for your version data**

Your class must inherit from `std_version`, which provides the common
properties (`scores_class`, `method_name`, `version_id`, `description`).
Add any method-specific properties:

```{r}
#| label: percentile-version-class
percentile_version <- S7::new_class(
  "percentile_version",
  parent = std_version,
  properties = list(
    ref_scores = S7::class_numeric
  ),
  constructor = function(
    scores_class,
    version_id,
    ref_scores,
    description = ""
  ) {
    S7::new_object(
      S7::S7_object(),
      scores_class = scores_class,
      method_name = "percentile",
      version_id = version_id,
      ref_scores = ref_scores,
      description = description
    )
  }
)
```

**Step 2: Write a registration function**

This is a thin wrapper around the internal `.register_std_version()`:

```{r}
#| label: register-percentile
register_percentile_version <- function(
  scores,
  version,
  ref_scores,
  description = "",
  overwrite = FALSE
) {
  scores_class <- S7::S7_class(scores)@name
  version_obj <- percentile_version(
    scores_class = scores_class,
    version_id = version,
    ref_scores = ref_scores,
    description = description
  )
  ntrs:::.register_std_version(version_obj, overwrite = overwrite)
}
```

**Step 3: Update the method to use `get_version_data()`**

Now redefine the method with a `version` parameter, and look up the
reference data from the registry:

```{r}
#| label: percentile-with-version
S7::method(std_using_percentile, npsych_scores) <- function(
  scores,
  ...,
  version
) {
  vdata <- get_version_data(scores, "percentile", version)
  raw <- as.numeric(remove_error_codes(scores))
  ecdf_fn <- stats::ecdf(vdata@ref_scores)
  ecdf_fn(raw) * 100
}
```

**Step 4: Register versions and set defaults**

```{r}
#| label: register-and-use
register_percentile_version(
  MOCATOTS(),
  version = "community",
  ref_scores = c(20, 22, 23, 24, 25, 25, 26, 27, 28, 29, 30),
  description = "Community-dwelling sample"
)

register_percentile_version(
  MOCATOTS(),
  version = "clinic",
  ref_scores = c(15, 17, 18, 20, 21, 22, 23, 24, 25, 26, 27),
  description = "Memory clinic sample"
)

set_std_defaults(
  MOCATOTS(),
  method = "percentile",
  version = "community",
  overwrite = TRUE
)
```

Now versions appear in the discovery functions:

```{r}
#| label: list-versions
list_method_versions(MOCATOTS(), "percentile")
```

And `std()` can resolve the default:

```{r}
#| label: std-with-version
moca <- MOCATOTS(c(22, 25, 28))

# Uses the default "community" version
std(moca)

# Or specify explicitly
std(moca, method = "percentile", version = "clinic")
```


## Part 2: Score-specific method overrides

You can register a method for a *specific* score class that takes precedence
over the parent class method. This is useful when certain tests need special
handling.

For example, Trail Making tests are timed — higher scores mean *worse*
performance. Let's make `std_using_percentile` flip the percentiles for
`TRAILA`:

```{r}
#| label: method-traila
S7::method(std_using_percentile, TRAILA) <- function(
  scores,
  ...,
  ref_scores
) {
  raw <- as.numeric(remove_error_codes(scores))
  ecdf_fn <- stats::ecdf(ref_scores)
  # Flip: 100 - percentile, so higher = better
  (1 - ecdf_fn(raw)) * 100
}
```

Now the dispatch works as expected:

```{r}
#| label: dispatch-demo
# MOCATOTS uses the parent method (higher = better)
std(
  MOCATOTS(c(22, 28)),
  method = "percentile",
  version = "community"
)

# TRAILA uses the specific method (lower time = better)
std(
  TRAILA(c(30, 60)),
  method = "percentile",
  ref_scores = c(20, 30, 40, 50, 60, 80, 100)
)
```

S7 dispatch order: specific class method → parent class method. The same
pattern is used in `ntrs` itself — `std_using_norms` and `std_using_regression`
have methods on `npsych_scores` (the parent) that handle all tests uniformly.


## Part 3: Packaging as an extension

The [ntrsTscores](https://github.com/rmtrane/ntrsTscores) package demonstrates
how to package a new standardization method as a standalone R package that
extends `ntrs`. It adds T-score standardization (`std_using_tscores`) for ~19
tests. Here's the key pattern:

### Package structure

```
ntrsTscores/
├── R/
│   ├── std_using_tscores.R            # S7 generic definition
│   ├── std_using_tscores-ANIMALS.R    # Method for ANIMALS
│   ├── std_using_tscores-TRAILA.R     # Method for TRAILA
│   ├── ...                            # One file per test
│   ├── scale_scores.R                 # Helper: raw → scaled scores
│   ├── SS.R                           # Reference data (scaled score tables)
│   └── zzz.R                          # .onLoad / .onAttach hooks
├── DESCRIPTION
└── tests/
```

### Key patterns

**1. Define the generic once** — in `R/std_using_tscores.R`:

```{r}
#| label: tscores-generic
#| eval: false
std_using_tscores <- S7::new_generic(
  "std_using_tscores",
  dispatch_args = "scores"
)
```

Note that this method doesn't use a `version` parameter — the coefficients are
hardcoded per test. This is a valid design when there's only one set of norms.

**2. Register S7 methods in `.onLoad()`** — in `R/zzz.R`:

```{r}
#| label: onload-hook
#| eval: false
.onLoad <- function(...) {
  S7::methods_register()
}
```

`S7::methods_register()` scans the package namespace and registers all
`S7::method()` calls. This is essential — without it, your methods won't be
found.

**3. Set defaults in `.onAttach()`** — so they're set when the user loads the
package:

```{r}
#| label: onattach-hook
#| eval: false
.onAttach <- function(...) {
  ntrs::set_std_defaults(REY123(), method = "tscores")
}
```

**4. Each method file** follows a consistent pattern — apply test-specific
regression coefficients to scaled scores:

```{r}
#| label: tscores-method-example
#| eval: false
S7::method(std_using_tscores, ANIMALS) <- function(
  scores,
  ...,
  age,
  sex,
  educ
) {
  ss <- scale_scores(scores)
  tscore <- round(
    50 + (ss - (10.69 + age * -0.007 + sex * -0.654 + educ * 0.213)) / 0.311,
    0
  )
  pmax(pmin(tscore, 100), 0)
}
```

### DESCRIPTION dependencies

Your extension package needs:

```{yaml}
Imports:
  ntrs, 
  S7
```

`Imports` is sufficient — `ntrs` generic discovery (`.find_std_generics()`)
searches all **loaded** namespaces, and `Imports` loads the `ntrs` namespace.
Users who want score constructors like `MOCATOTS()` on the search path simply
call `library(ntrs)` themselves (which they likely already do).

If you prefer that `library(yourpkg)` automatically attaches `ntrs` for user
convenience, use `Depends: ntrs` instead.


## Summary

To add a new standardization method to `ntrs`:

1. **Create an S7 generic** named `std_using_<method>` dispatching on `scores`
2. **Implement methods** on `npsych_scores` (for all tests) or specific
   subclasses (for test-specific logic)
3. **Optionally** use the version registry if your method supports multiple
   parameter sets
4. **To package**: put the generic + methods in an R package, call
   `S7::methods_register()` in `.onLoad()`, and `Imports: ntrs` in DESCRIPTION
