---
title: "New Scores and Versions"
vignette: >
  %\VignetteIndexEntry{New Scores and Versions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ntrs)
```

This vignette shows how to extend `ntrs` by creating new neuropsychological
test score types and registering standardization versions for them. You might do
this to:

- Add a test that `ntrs` doesn't ship with
- Register your own normative data for an existing test
- Use site-specific regression coefficients

## Creating a new score type

Every test score in `ntrs` is an S7 subclass of `npsych_scores`, created with
the `new_npsych_scores()` factory function. Let's create one for a fictional
"Digit Symbol" test:

```{r}
#| label: create-dsym
DSYM <- new_npsych_scores(
  name = "DSYM",
  label = "Digit Symbol",
  domain = "Processing Speed",
  short_descriptor = "Digit Symbol Substitution — total correct in 90s",
  range = c(0, 80),
  codes = c(
    "Not administered" = 88,
    "Could not complete" = 99
  )
)
```

The arguments are:

- **`name`**: The class name (used as the registry key and in output)
- **`label`**: A human-readable short name
- **`domain`**: The cognitive domain (optional)
- **`short_descriptor`**: A longer description (optional)
- **`range`**: The minimum and maximum *valid* scores (inclusive)
- **`codes`**: Named numeric vector of special/error codes that may fall outside the
  valid range

`DSYM` is now a constructor function. Use it to wrap raw numeric data:

```{r}
#| label: dsym-construct
scores <- DSYM(c(45, 62, 88, 33, NA))
scores
```

The value `88` is stored as-is because it's a registered code (\"Not
administered\"). Validation catches values outside the range *and* codes:

```{r}
#| label: dsym-validation-error
#| error: true
DSYM(c(45, 100))
```

The new score is immediately discoverable:

```{r}
#| label: dsym-discoverable
"DSYM" %in% list_npsych_scores()
```

### Working with error codes

Use `remove_error_codes()` to replace codes with `NA` for analysis, or
`replace_codes()` to see their labels:

```{r}
#| label: error-codes
remove_error_codes(scores)

replace_codes(scores)
```


## Registering a norms version

To standardize scores, you need to register at least one standardization
version. Let's start with norms-based standardization, which uses a lookup table
of means and standard deviations stratified by covariates. This is implemented in the `ntrs` method `std_using_norms()`, so we just need to provide the necessary data and register it with `register_norms_version()`, and the method will be available for our new `DSYM` scores.

### The lookup table

The lookup table is a data frame with columns for each covariate plus `m`
(mean) and `sd` (standard deviation). Optionally, it can include `n` (sample
size). Here's a small example stratified by age group and sex:

```{r}
#| label: norms-lookup-table
dsym_norms <- data.frame(
  age = factor(rep(c("<65", "65-74", "75+"), each = 2)),
  sex = factor(rep(c("m", "f"), 3)),
  m = c(55, 58, 48, 51, 40, 43),
  sd = c(10, 9, 11, 10, 12, 11)
)

dsym_norms
```

### Covariate functions

Since users pass raw covariates (e.g., `age = 72`, `sex = 1`), you need
**covariate functions** that transform them to match the lookup table format.
Each function takes a numeric vector and returns the transformed values:

```{r}
#| label: norms-covar-fns
dsym_covar_fns <- list(
  age = \(x) {
    dplyr::case_when(
      x < 65 ~ "<65",
      x >= 65 & x < 75 ~ "65-74",
      x >= 75 ~ "75+"
    ) |>
      factor(levels = c("<65", "65-74", "75+"))
  },
  sex = \(x) factor(x, levels = c(1, 2), labels = c("m", "f"))
)
```

The names in `covar_fns` must match the covariate column names in the lookup
table (here, `age` → `age`, `sex` → `sex`). 

### Registering

Now register with `register_norms_version()`:

```{r}
#| label: register-norms
register_norms_version(
  scores = DSYM(),
  version = "example_v1",
  lookup_table = dsym_norms,
  covar_fns = dsym_covar_fns
)
```

The version is now discoverable:

```{r}
#| label: list-norms-versions
list_method_versions(DSYM(), "norms")
```

We can standardize using it:

```{r}
#| label: std-norms
std(
  DSYM(c(45, 62, 33)),
  method = "norms",
  version = "example_v1",
  age = c(72, 68, 80),
  sex = c(1, 2, 1)
)
```

## Registering a regression version

Regression-based standardization uses a linear model:
$z = (\text{raw} - \text{predicted}) / \text{RMSE}$, where predicted values come
from regression coefficients.

### The coefficients

Provide a named numeric vector with `intercept`, `rmse`, and any covariate
names:

```{r}
#| label: regression-coefs
dsym_coefs <- c(
  intercept = 65.2,
  age = -0.35,
  sex = 3.1,
  educ = 0.8,
  rmse = 9.5
)
```

### Covariate functions (optional)

For regression, `covar_fns` transform raw inputs before they enter the linear
model. If omitted, identity functions are used for each covariate. Here we
recode sex from `1/2` to `0/1`, and truncate education at 8 and 30 years:

```{r}
#| label: register-regression
register_regression_version(
  scores = DSYM(),
  version = "example_reg",
  coefs = dsym_coefs,
  covar_fns = list(
    age = identity,
    sex = \(x) as.numeric(x == 2),
    educ = \(x) pmin(pmax(x, 8), 30)
  )
)
```

Again, the version is discoverable and can be used for standardization:

```{r}
#| label: list-regression-versions
list_method_versions(DSYM(), "regression")

std(
  DSYM(c(45, 62, 33)),
  method = "regression",
  version = "example_reg",
  age = c(72, 68, 80),
  sex = c(1, 2, 1),
  educ = c(16, 12, 14)
)
```


We can verify that the education variable is indeed truncated by comparing the result at the boundary to a value above it:

```{r}
#| label: verify-educ-truncation
std(
  DSYM(c(45, 45)),
  method = "regression",
  version = "example_reg",
  age = c(72, 72),
  sex = c(1, 1),
  educ = c(30, 40) # 40 is truncated to
)
```

## Setting defaults and standardizing

Set a default method and version so users can call `std()` without specifying
them:

```{r}
#| label: set-defaults
set_std_defaults(DSYM(), method = "regression", version = "example_reg")

get_std_defaults(DSYM())
```

Now standardize:

```{r}
#| label: std-with-defaults
test_scores <- DSYM(c(45, 62, 33))

# Uses the default (regression)
std(test_scores, age = 72, sex = 1, educ = 16)

# Explicitly use norms
std(test_scores, method = "norms", version = "example_v1", age = 72, sex = 1)
```

## Optional: `raw_scores_fn` and `post_proc_fn`

Both `register_norms_version()` and `register_regression_version()` accept two
optional transformation functions:

- **`raw_scores_fn`**: Applied to raw scores *before* standardization. Useful
  for log-transforms or reversals.
- **`post_proc_fn`**: Applied to z-scores *after* standardization. Useful for
  sign flips (e.g., when higher raw scores mean worse performance).

For example, Trail Making tests are timed, so higher scores = worse
performance. A `post_proc_fn = \(x) -x` flips the z-scores so that positive
always means better:

```{r}
#| label: post-proc-example
#| eval: false
register_norms_version(
  scores = TRAILA(),
  version = "my_flipped",
  lookup_table = my_traila_norms,
  covar_fns = my_covar_fns,
  post_proc_fn = \(x) -x
)
```


## Adding a version to an existing score

You aren't limited to new scores — you can register additional versions for any
score that `ntrs` already provides. For instance, to add site-specific norms for
`MOCATOTS`:

```{r}
#| label: add-version-existing-score
my_moca_norms <- data.frame(
  age = factor(c("<70", "70+", "<70", "70+")),
  sex = factor(c("m", "m", "f", "f")),
  m = c(26.1, 24.3, 26.5, 24.8),
  sd = c(2.8, 3.1, 2.6, 2.9)
)

register_norms_version(
  scores = MOCATOTS(),
  version = "my_site",
  lookup_table = my_moca_norms,
  covar_fns = list(
    age = \(x) factor(ifelse(x < 70, "<70", "70+")),
    sex = \(x) factor(x, levels = c(1, 2), labels = c("m", "f"))
  )
)

list_method_versions(MOCATOTS(), "norms")
```

The new version coexists with the built-in ones and can be used via
`std(moca_scores, method = "norms", version = "my_site", age = 72, sex = 1)`.


## Packaging it up

If you want your new scores and versions to be available every time you load
your package, follow the pattern used by `ntrs` internally:

1. Create `R/npsych_scores-DSYM.R` with the `new_npsych_scores()` call and a
   `.setup_DSYM_versions()` function that registers all versions and sets
   defaults.

2. In your package's `.onLoad()`, call `S7::methods_register()` and then your
   setup functions. If your package extends `ntrs`, the `.onLoad()` hook in
   `ntrs` will auto-discover any `.setup_*_versions()` functions in the `ntrs`
   namespace, but for your own package you'll need to call them explicitly. It 
   is recommended that you do this in `.onAttach()`.

See `R/npsych_scores-MOCATOTS.R` in the `ntrs` source for a complete template.
