---
title: "Getting Started"
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
suppressPackageStartupMessages(
  library(NpsychBatteryNormsS3)
)
```

## Overview

NpsychBatteryNormsS3 provides an S3-based interface for standardizing
neuropsychological test scores. The typical workflow is:

1. Create score objects using factory functions (e.g., `MOCATOTS()`, `ANIMALS()`)
2. Standardize scores with `std()` or batch-standardize a data frame with `std_data()`

## Creating score objects

Each neuropsychological test has a dedicated factory function that wraps raw
numeric scores with metadata (valid range, error codes, etc.):

```{r}
moca <- MOCATOTS(c(25, 28, 30))
moca
```

To see all available tests:

```{r}
list_npsych_scores()
```

## Discovering standardization options

Before standardizing, you can check which methods and versions are available
for a given test:
```{r}
list_std_methods(moca)

list_method_versions(moca, "norms")

list_method_versions(moca, "regression")
```

Each test has registered defaults so you don't need to specify a method every
time:

```{r}
get_std_defaults(moca)
```

These can be changed globally with `set_std_defaults()`:

```{r}
set_std_defaults(
  moca,
  method = "norms",
  version = "nacc",
  overwrite = TRUE
)

get_std_defaults(moca)
```

Do note that changes to the default will not be saved from session to session, so you may want to set them in your `.Rprofile` if you want them to persist.

## Standardizing a single score vector

Use `std()` to convert raw scores to z-scores. Covariates (age, sex,
education, etc.) are passed as named arguments:

```{r}
z <- std(moca, age = 72, sex = 1, educ = 16)
z

set_std_defaults(
  moca,
  method = "regression",
  version = "updated_2025.06",
  overwrite = TRUE
)

z_reg <- std(moca, age = 72, sex = 1, educ = 16, race = 1)
z_reg
```

To override the default method or version, specify them explicitly:

```{r}
z_norms <- std(
  moca,
  method = "norms",
  version = "updated",
  age = 72,
  sex = 1,
  educ = 16
)
z_norms
```

## Batch standardization with `std_data()`

When you have a data frame with multiple test score columns, `std_data()`
standardizes all of them at once:

```{r}
df <- data.frame(
  age = c(72, 75, 68),
  sex = c(1, 2, 1),
  educ = c(16, 14, 12),
  race = c(1, 2, 1),
  moca = MOCATOTS(c(25, 28, 22)),
  animals = ANIMALS(c(18, 20, 16))
)

result <- std_data(
  df,
  age = age,
  sex = sex,
  educ = educ,
  race = race
)

result
```

You can override the method/version per test class:

```{r}
result2 <- std_data(
  df,
  methods = list(
    MOCATOTS = c(method = "norms", version = "nacc")
  ),
  age = age,
  sex = sex,
  educ = educ,
  race = race
)

result2
```

## Changing defaults

If you prefer a different default method for your session, use
`set_std_defaults()`:

```{r}
#| eval: false
set_std_defaults(MOCATOTS(), method = "norms", version = "nacc")

# Subsequent calls use the new default
std(MOCATOTS(c(25, 28)), age = 72, sex = 1, educ = 16)
```
