---
title: "Working with NACC Data"
vignette: >
  %\VignetteIndexEntry{Working with NACC Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ntrs)
```

`ntrs` is designed around the [National Alzheimer's Coordinating Center
(NACC)](https://naccdata.org/) Uniform Data Set (UDS). This vignette walks
through a realistic workflow: starting from raw NACC-style data, computing
derived scores, and batch-standardizing everything.

## The demo data

`ntrs` ships with a synthetic dataset, `demo_data`, that mimics the structure
of a NACC data extract. It contains demographic variables, cognitive test
scores, and functional assessment items — all with realistic error codes.

```{r}
#| label: demo-data
dim(demo_data)
names(demo_data)[1:20]
```

The column names match NACC variable names (e.g., `MOCATOTS`, `TRAILA`,
`ANIMALS`). You can look up any variable in the Researcher's Data Dictionary
via the `rdd` object:

```{r}
#| label: rdd-lookup
rdd$MOCATOTS$short_descriptor
rdd$MOCATOTS$range
rdd$MOCATOTS$codes
```

## Step 1: Convert columns to `npsych_scores`

Raw NACC data arrives as plain numeric columns. The first step is to wrap the
columns you want to standardize as `npsych_scores` objects. This adds
validation, range checking, and error-code awareness.

```{r}
#| label: convert-scores
df <- demo_data

df$MOCATOTS <- MOCATOTS(df$MOCATOTS)
df$ANIMALS <- ANIMALS(df$ANIMALS)
df$TRAILA <- TRAILA(df$TRAILA)
df$CRAFTVRS <- CRAFTVRS(df$CRAFTVRS)
df$UDSBENTD <- UDSBENTD(df$UDSBENTD)
```

Once converted, the columns carry metadata:

```{r}
#| label: inspect-scores
df$MOCATOTS[1:10]
```

## Step 2: Compute derived scores

Some tests in the NACC battery don't appear directly in the data — they must
be calculated from component variables. `ntrs` provides `calc_*` helper
functions for these.

### MoCA Clock Drawing total

The MoCA clock drawing score is the sum of three sub-items (contour, numbers,
hands), each scored 0 or 1:

```{r}
#| label: calc-mocaclock
df$MOCACLOCK <- calc_MOCACLOCK(
  MOCACLOC = demo_data$MOCACLOC,
  MOCACLON = demo_data$MOCACLON,
  MOCACLOH = demo_data$MOCACLOH
)

df$MOCACLOCK[1:10]
```

### RAVLT Total Learning

The Rey Auditory Verbal Learning Test (RAVLT) total learning score is the sum
of trials 1 through 5:

```{r}
#| label: calc-reytotal
df$REYTOTAL <- calc_REYTOTAL(
  rey1rec = REY1REC(demo_data$REY1REC),
  rey2rec = REY2REC(demo_data$REY2REC),
  rey3rec = REY3REC(demo_data$REY3REC),
  rey4rec = REY4REC(demo_data$REY4REC),
  rey5rec = REY5REC(demo_data$REY5REC)
)

df$REYTOTAL[1:10]
```

### RAVLT Recognition Accuracy

Recognition accuracy combines true positives (`REYTCOR`) and false positives
(`REYFPOS`):

```{r}
#| label: calc-reyarec
df$REYAREC <- calc_REYAREC(
  reytcor = REYTCOR(demo_data$REYTCOR),
  reyfpos = REYFPOS(demo_data$REYFPOS)
)

df$REYAREC[1:10]
```

### Functional Activities Score

The Functional Activities Score (FAS) summarizes ten everyday-functioning
items. Error codes like 8 ("Not applicable") are treated as 0:

```{r}
#| label: calc-fas
df$FAS <- calc_FAS(
  BILLS = demo_data$BILLS,
  TAXES = demo_data$TAXES,
  SHOPPING = demo_data$SHOPPING,
  GAMES = demo_data$GAMES,
  STOVE = demo_data$STOVE,
  MEALPREP = demo_data$MEALPREP,
  EVENTS = demo_data$EVENTS,
  PAYATTN = demo_data$PAYATTN,
  REMDATES = demo_data$REMDATES,
  TRAVEL = demo_data$TRAVEL
)

head(df$FAS)
```

## Step 3: Batch standardization with `std_data()`

Now that our data frame contains `npsych_scores` columns, we can standardize
them all at once. Covariates are passed as bare column names referencing
columns in the data frame:

```{r}
#| label: std-data
result <- std_data(
  df,
  age = NACCAGE,
  sex = SEX,
  educ = EDUC,
  race = RACE
)
```

`std_data()` finds every `npsych_scores` column, standardizes it using the
registered default method and version, and adds a new column with a `z_`
prefix:

```{r}
#| label: inspect-result
# Show the z-score columns
z_cols <- grep("^z_", names(result), value = TRUE)
z_cols

head(result[, c("MOCATOTS", "z_MOCATOTS", "ANIMALS", "z_ANIMALS")])
```

### Tracking which method was used

The result carries metadata attributes. Use `methods_from_std_data()` to see
which method and version were applied to each score:

```{r}
#| label: methods-from-std-data
methods_from_std_data(result)
```

### Keeping raw scores alongside z-scores

By default, the original `npsych_scores` columns are left in place. If you
prefer to rename them with a prefix (e.g., `raw_`) for clearer bookkeeping,
use the `prefix_raw` argument:

```{r}
#| label: prefix-raw
result2 <- std_data(
  df,
  prefix_raw = "raw_",
  age = NACCAGE,
  sex = SEX,
  educ = EDUC,
  race = RACE
)

# Now we have raw_MOCATOTS and z_MOCATOTS
names(result2)[grep("MOCATOTS", names(result2))]
```

### Overriding defaults per test

You can specify different methods or versions for specific tests via the
`methods` argument, keyed by class name:

```{r}
#| label: override-methods
result3 <- std_data(
  df,
  methods = list(
    MOCATOTS = list(method = "norms", version = "nacc"),
    ANIMALS = list(method = "norms", version = "updated")
  ),
  age = NACCAGE,
  sex = SEX,
  educ = EDUC,
  race = RACE
)
```

### Standardizing a subset of columns

If you only want to standardize specific columns, use `.cols`:

```{r}
#| label: subset-cols
result4 <- std_data(
  df,
  .cols = c("MOCATOTS", "ANIMALS", "TRAILA"),
  age = NACCAGE,
  sex = SEX,
  educ = EDUC,
  race = RACE
)

grep("^z_", names(result4), value = TRUE)
```

## Putting it all together

For simplicity, consider just a subset of the `demo_data`.

```{r}
demo_subset <- demo_data[, c(
  "NACCID",
  "NACCAGE",
  "SEX",
  "EDUC",
  "RACE",
  "MOCATOTS",
  "ANIMALS",
  "TRAILA",
  "REY1REC",
  "REY2REC",
  "REY3REC",
  "REY4REC",
  "REY5REC",
  "MOCACLOC",
  "MOCACLON",
  "MOCACLOH",
  "REYTCOR",
  "REYFPOS",
  "BILLS",
  "TAXES",
  "SHOPPING",
  "GAMES",
  "STOVE",
  "MEALPREP",
  "EVENTS",
  "PAYATTN",
  "REMDATES",
  "TRAVEL"
)]
``` 

Here is the full workflow, from raw data to standardized scores:

```{r}
# 1. Start with raw NACC data
df <- demo_subset
# 2. Convert test columns to npsych_scores
df$MOCATOTS <- MOCATOTS(df$MOCATOTS)
df$ANIMALS <- ANIMALS(df$ANIMALS)
df$TRAILA <- TRAILA(df$TRAILA)
# ... repeat for each test of interest

# 3. Compute derived scores
df$REYTOTAL <- calc_REYTOTAL(
  REY1REC(df$REY1REC),
  REY2REC(df$REY2REC),
  REY3REC(df$REY3REC),
  REY4REC(df$REY4REC),
  REY5REC(df$REY5REC)
)
df$MOCACLOCK <- calc_MOCACLOCK(df$MOCACLOC, df$MOCACLON, df$MOCACLOH)
df$FAS <- calc_FAS(
  BILLS = df$BILLS,
  TAXES = df$TAXES,
  SHOPPING = df$SHOPPING,
  GAMES = df$GAMES,
  STOVE = df$STOVE,
  MEALPREP = df$MEALPREP,
  EVENTS = df$EVENTS,
  PAYATTN = df$PAYATTN,
  REMDATES = df$REMDATES,
  TRAVEL = df$TRAVEL
)

# 4. Standardize
result <- std_data(
  df,
  age = NACCAGE,
  sex = SEX,
  educ = EDUC,
  race = RACE
)
```

We can similarly work within the `tidyverse` if that's your preference:

```{r}
library(dplyr)

df_tidy <- demo_subset |>
  mutate(
    MOCATOTS = MOCATOTS(MOCATOTS),
    ANIMALS = ANIMALS(ANIMALS),
    TRAILA = TRAILA(TRAILA),
    REYTOTAL = calc_REYTOTAL(
      REY1REC(REY1REC),
      REY2REC(REY2REC),
      REY3REC(REY3REC),
      REY4REC(REY4REC),
      REY5REC(REY5REC)
    ),
    MOCACLOCK = calc_MOCACLOCK(MOCACLOC, MOCACLON, MOCACLOH),
    FAS = calc_FAS(
      BILLS,
      TAXES,
      SHOPPING,
      GAMES,
      STOVE,
      MEALPREP,
      EVENTS,
      PAYATTN,
      REMDATES,
      TRAVEL
    )
  )

df_tidy_result <- std_data(
  df_tidy,
  age = NACCAGE,
  sex = SEX,
  educ = EDUC,
  race = RACE
)
```

Or `data.table`:

```{r}
library(data.table)

df_dt <- as.data.table(demo_subset)

df_dt[,
  c(
    "MOCATOTS",
    "ANIMALS",
    "TRAILA",
    "REYTOTAL",
    "MOCACLOCK",
    "FAS"
  ) := list(
    MOCATOTS(MOCATOTS),
    ANIMALS(ANIMALS),
    TRAILA(TRAILA),
    calc_REYTOTAL(
      REY1REC(REY1REC),
      REY2REC(REY2REC),
      REY3REC(REY3REC),
      REY4REC(REY4REC),
      REY5REC(REY5REC)
    ),
    calc_MOCACLOCK(MOCACLOC, MOCACLON, MOCACLOH),
    calc_FAS(
      BILLS,
      TAXES,
      SHOPPING,
      GAMES,
      STOVE,
      MEALPREP,
      EVENTS,
      PAYATTN,
      REMDATES,
      TRAVEL
    )
  )
]

df_dt_result <- std_data(
  df_dt,
  age = NACCAGE,
  sex = SEX,
  educ = EDUC,
  race = RACE
)
```


We can verify that all three approaches yield the same result (par differences in classes):

```{r}
waldo::compare(as.data.frame(result), as.data.frame(df_tidy_result))
waldo::compare(as.data.frame(df_tidy_result), as.data.frame(df_dt_result))
```